\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-30 paper.qmd
Rearrange some figures and removed unnecessary images and unnessary sections

@@

num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 72, aic = -21882.462)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "UMAP", tot_bins = 98, aic = -20013.266)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PHATE", tot_bins = 90, aic = -15036.91)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 154, aic = -23001.90)

num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
  tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 132, aic = -21330.059)

# knitr::kable(
#   num_bins_for_nldr_methods_df,
#   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
#              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# ) |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE)
num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins', "AIC")
num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)

xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
  xtable2kable() |> 
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
  row_spec(0, bold = TRUE) |>
  row_spec(4, bold=T, color = "red")

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Total number of bins")

@@

## To draw with AIC
aic_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Benchmark value")
## Effective number of bins along x-axis

mse_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Benchmark value")

@@

## To draw with AIC
aic_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Benchmark value")
## Effective number of bins along x-axis

mse_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Benchmark value")

@@

```{r}
#| echo: false
#| fig-cap: "2D layouts from UMAP applied for the S-curve data: (a) UMAP (n_neighbors = 7), (b) UMAP (n_neighbors = 15), (c) UMAP (n_neighbors = 32), (d) UMAP (n_neighbors = 50). Is there a best parameter choice in representing UMAP or are they all providing  equivalent information?"
#| label: fig-nldervisUMAP
#| out-width: 100%

@@

aic_plot_param + mse_plot_param_training + mse_plot_param_test  +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

@@

### Model selection {#sec-selection}

In the process of selecting the most suitable NLDR technique to represent high-D data, the Akaike Information Criterion (AIC) stands as a crucial tool for insightful comparisons. Once the effective number of bins has been determined for each method, AIC values can be compared to identify the best NLDR technique with the lowest AIC value. For example, as shown in @fig-diagnosticpltScurve (a), TriMAP is the most suitable NLDR technique to represent the S-curve data (see table 1).

>>>>

```{r}
#| echo: false
#| fig-cap: "2D layouts from UMAP applied for the S-curve data: (a) UMAP (n_neighbors = 7), (b) UMAP (n_neighbors = 15), (c) UMAP (n_neighbors = 32), (d) UMAP (n_neighbors = 50). Is there a best hyperparameter choice in representing UMAP or are they all providing  equivalent information?"
#| label: fig-nldervisUMAP
#| out-width: 100%

@@

# num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 72, aic = -21882.462)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "UMAP", tot_bins = 98, aic = -20013.266)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "PHATE", tot_bins = 90, aic = -15036.91)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 154, aic = -23001.90)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 132, aic = -21330.059)
# 
# # knitr::kable(
# #   num_bins_for_nldr_methods_df,
# #   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
# #              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# # ) |> 
# #   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
# #   row_spec(0, bold = TRUE)
# num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
# colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins', "AIC")
# num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)
# 
# xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
#   xtable2kable() |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE) |>
#   row_spec(4, bold=T, color = "red")

@@

## To draw with AIC
aic_plot <- ggplot(MSE_df |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Total number of bins")

@@

## To draw with AIC
aic_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Benchmark value")
## Effective number of bins along x-axis

mse_plot_lwd <- ggplot(MSE_df_lwd |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_hex,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Benchmark value")

@@

## To draw with AIC
aic_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("Benchmark value")
## Effective number of bins along x-axis

mse_plot_lg <- ggplot(error_df_lg |> dplyr::filter(method == "UMAP"), aes(x = benchmark_rm_lg,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("MSE") +
  xlab("Benchmark value")

@@

<!--
```{r}
#| warning: false
#| echo: false

@@

aic_plot_param + mse_plot_param_training + mse_plot_param_test  +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 3) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```
-->

@@

<!--### Model selection {#sec-selection}

In the process of selecting the most suitable NLDR technique to represent high-D data, the Akaike Information Criterion (AIC) stands as a crucial tool for insightful comparisons. Once the effective number of bins has been determined for each method, AIC values can be compared to identify the best NLDR technique with the lowest AIC value. For example, as shown in @fig-diagnosticpltScurve (a), TriMAP is the most suitable NLDR technique to represent the S-curve data (see table 1). -->

### 2024-01-30 paper.tex
Rearrange some figures and removed unnecessary images and unnessary sections

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

@@

\begin{table}[H]
\centering
\begin{tabular}{rllr}
  \hline
\textbf{} & \textbf{NLDR technique} & \textbf{total number of bins} & \textbf{AIC }\\ 
  \hline
\cellcolor{gray!6}{1} & \cellcolor{gray!6}{tSNE} & \cellcolor{gray!6}{72} & \cellcolor{gray!6}{-21882.46 }\\ 
  2 & UMAP & 98 & -20013.27 \\ 
\cellcolor{gray!6}{  3} & \cellcolor{gray!6}{PHATE} & \cellcolor{gray!6}{90} & \cellcolor{gray!6}{-15036.91 }\\ 
\textcolor{red}{\textbf{  4}} & \textcolor{red}{\textbf{TriMAP}} & \textcolor{red}{\textbf{154}} & \textcolor{red}{\textbf{-23001.90 }}\\ 
\cellcolor{gray!6}{  5} & \cellcolor{gray!6}{PaCMAP} & \cellcolor{gray!6}{132} & \cellcolor{gray!6}{-21330.06 }\\ 
   \hline
\end{tabular}
\caption{Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.} 
\label{tbl-msebinsnldr}
\end{table}

@@

\caption{\label{fig-nldervisUMAP}2D layouts from UMAP applied for the
S-curve data: (a) UMAP (n\_neighbors = 7), (b) UMAP (n\_neighbors = 15),
(c) UMAP (n\_neighbors = 32), (d) UMAP (n\_neighbors = 50). Is there a
best parameter choice in representing UMAP or are they all providing
equivalent information?}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltDiffParam-1.pdf}

\caption{\label{fig-diagnosticpltDiffParam}Goodness of fit statistics
from different n\_neighbors parameter of UMAP applied to training
S-curve dataset. What is the best parameter choice in UMAP to represent
the original data in 2D?}

@@

\hypertarget{sec-selection}{%
\subsection{Model selection}\label{sec-selection}}

In the process of selecting the most suitable NLDR technique to
represent high-D data, the Akaike Information Criterion (AIC) stands as
a crucial tool for insightful comparisons. Once the effective number of
bins has been determined for each method, AIC values can be compared to
identify the best NLDR technique with the lowest AIC value. For example,
as shown in Figure~\ref{fig-diagnosticpltScurve} (a), TriMAP is the most
suitable NLDR technique to represent the S-curve data (see table 1).

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, breakable, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\caption{\label{fig-nldervisUMAP}2D layouts from UMAP applied for the
S-curve data: (a) UMAP (n\_neighbors = 7), (b) UMAP (n\_neighbors = 15),
(c) UMAP (n\_neighbors = 32), (d) UMAP (n\_neighbors = 50). Is there a
best hyperparameter choice in representing UMAP or are they all
providing equivalent information?}

@@

@@

@@
